Dominance Regions for Rank Two Cluster Algebras

### 2022-04-06 dominance_regions.tex
Small notation change, I like it and I hate it, thoughts?

  Fix integers $b,c>0$.  The cluster algebra $\cA(b,c)$ is the $\ZZ$-subalgebra of $\QQ(x_1,x_2)$ generated by \emph{cluster variables} $x_m$, $m\in\ZZ$, defined recursively by
  \begin{equation}
    x_{m-1}x_{m+1}=\begin{cases} x_m^b+1 & \text{if $m$ is odd;}\\ x_m^c+1 & \text{if $m$ is even.} \end{cases}
  \end{equation}
  By the Laurent Phenomenon \cite{fomin-zelevinsky}, each $x_m$ is actually an element of $\ZZ[x_k^{\pm1},x_{k+1}^{\pm1}]$ for any $k\in\ZZ$.
  Moreover, these Laurent polynomials are known to have positive coefficients \cite{LLZ,lee-schiffler,GHKK}.

  Our goal in this note is to explicitly compute the possible bases which are pointed with respect to the embedding $\cA(b,c)\into\ZZ[x_k^{\pm1},x_{k+1}^{\pm1}]$ for every $k\in\ZZ$.
  That is, a basis element with \emph{$\bfg$-vector} $\lambda\in\ZZ^2$ when expanded in terms of $\{x_1,x_2\}$ can be written in the form
  \begin{equation}
    \label{eq:pointed}
    x_1^{\lambda_1}x_2^{\lambda_2}\sum\limits_{a_1,a_2 \ge 0} c^\lambda_{a_1,a_2} x_1^{-ba_1} x_2^{ca_2}
  \end{equation}
  with $c^\lambda_{0,0}=1$.
  Being \emph{pointed} means an analogous structure reproduces after expanding in terms of any pair $\{x_k,x_{k+1}\}$.
  %The collection of all $\bfg$-vectors $\lambda\in\ZZ^2$ parametrize the elements of any pointed basis.

  Important examples of pointed elements are the \emph{cluster monomials} of $\cA(b,c)$, i.e. the elements that can be written in the form $x_k^{a_k}x_{k+1}^{a_{k+1}}$ for some $k\in\ZZ$ and $a_k,a_{k+1}$ non-negative integers.
  Indeed, by [someresult of Qin] cluster monomials are part of any pointed basis in any cluster algebra.
  \say{Check if this needs full rank}

@@

  A number of bases for rank two cluster algebras with various defining properties have been constructed.
  A natural starting basis is the so-called \emph{greedy basis} in which all $c^\lambda_{a_1,a_2}$ are non-negative and these coefficients are chosen to be minimal.
  It is known that this choice coincides with the theta basis \cite{GHKK, CGMMRSW}.
  The \emph{standard monomial bases} of $\cA(b,c)$ come from natural identifications with the subalgebras of $\ZZ[x_1^{\pm1},x_2^{\pm1}]$ generated by $x_{k-1}$, $x_k$, $x_{k+1}$, and $x_{k+2}$ for $k\in\ZZ$.
  However, these bases depend on $k$ and the triangular basis is defined by a triangular transition with the stipulation that the basis does not depend on the choice of starting cluster.
  It would be useful to explicitly (or combinatorially) compute the transitions between these bases.
  However, explicit calculations reveal that the $\bfg$-vectors actually appearing in these transitions are far smaller than the entire dominance polygon.

@@

  Given a Laurent polynomial in $\ZZ[x_1^{\pm1},x_2^{\pm1}]$, its \emph{support} is the set of its exponent vectors inside $\ZZ^2$.
  \begin{theorem}
    \label{th:maximum support}
    For a $\bfg$-vector $\lambda=(\lambda_1,\lambda_2)\in\ZZ^2$ the support for a pointed basis vector $x_\lambda$ is contained in the (possibly degenerate) lattice quadrilateral $\cS_\lambda$ given as follows:
    \begin{enumerate}
      \item If $\lambda$ lies outside of $\cI$, there are the following possibilities:
        \begin{enumerate}
          \item If $0\le\lambda_1,\lambda_2$, then $\cS_\lambda$ is just the point $\lambda$.
          \item If $\lambda_2 < 0$ and $0\le\lambda_1+b\lambda_2$, then $\cS_\lambda$ is the segment joining $\lambda$ and $(\lambda_1+b\lambda_2,\lambda_2)$.
          \item If $\lambda_1 < 0$ and $0\le\lambda_2$, then $\cS_\lambda$ is the segment joining $\lambda$ and $(\lambda_1,-c\lambda_1+\lambda_2)$.
          \item If $\lambda_1,\lambda_2 < 0$, then $\cS_\lambda$ has corner vertices $\lambda$, $(\lambda_1+b\lambda_2,\lambda_2)$, $\big(\lambda_1+b\lambda_2,-c\lambda_1-(bc-1)\lambda_2\big)$, and $(\lambda_1,-c\lambda_1+\lambda_2)$.
          \item If $\lambda_1 > 0$, $\lambda_1+b\lambda_2 < 0$, and $-c\lambda_1-(bc-1)\lambda_2\le 0$, then $\cS_\lambda$ has corner vertices $\lambda$, $(\lambda_1+b\lambda_2,\lambda_2)$, $\big(\lambda_1+b\lambda_2,-c\lambda_1-(bc-1)\lambda_2\big)$, and $\big((bc+1)\lambda_1-b^2c\lambda_2,-c\lambda_1-(bc-1)\lambda_2\big)$.
          \item If $\lambda_1 > 0$, $\lambda_1+b\lambda_2 < 0$, and $0 < -c\lambda_1-(bc-1)\lambda_2$, then $\cS_\lambda$ has corner vertices $\lambda$, $(\lambda_1+b\lambda_2,\lambda_2)$, $\big(\lambda_1+b\lambda_2,-c\lambda_1-(bc-1)\lambda_2\big)$, and $(0,0)$.
        \end{enumerate}
      \item If $\lambda$ lies inside of $\cI$, then $\cS_\lambda$ has corner vertices $\lambda$, $(\lambda_1+b\lambda_2,\lambda_2)$, $\big(\lambda_1+b\lambda_2,-c\lambda_1-(bc-1)\lambda_2\big)$, and $\frac{bc+\sqrt{bc(bc-4)}}{2\sqrt{bc(bc-4)}}\big(2\lambda_1+b\lambda_2,-c\lambda_1-(bc-2)\lambda_2\big)$.
    \end{enumerate}
  \end{theorem}

@@

\section{$\bfg$-vector Mutations}
  We begin by studying transformations of $\RR^2$ which determine the change of $\bfg$-vectors when expanding a pointed expression of the form~\eqref{eq:pointed} in terms of another cluster.
  Write $\phi_0:\RR^2\to\RR^2$ for the identity map and define piecewise-linear maps $\phi_{\pm 1}:\RR^2\to\RR^2$ as follows:
  \begin{equation}
    \phi_1(\lambda)
    =
    \begin{cases} 
      (-\lambda_1,c\lambda_1+\lambda_2) & \text{if $\lambda_1 \ge 0$;}\\
      (-\lambda_1,\lambda_2) & \text{if $\lambda_1 < 0$;}
    \end{cases}
    \qquad
    \phi_{-1}(\lambda)
    =
    \begin{cases} 
      (\lambda_1,-\lambda_2) & \text{if $\lambda_2 \ge 0$;}\\
      (\lambda_1+b\lambda_2,-\lambda_2) & \text{if $\lambda_2 < 0$.}
    \end{cases}
  \end{equation}
  For $k\in\ZZ$ with $|k|>1$, define piecewise-linear maps
  \[\phi_k
    :=
    \begin{cases}
      (\phi_{-1}^{-1}\phi_1)^j & \text{if $k=2j$;}\\
      \phi_1(\phi_{-1}^{-1}\phi_1)^j & \text{if $k=2j+1$.}\\
    \end{cases}
  \]
  These determine the dominance region as follows.
  \begin{definition}
    For $\lambda\in\ZZ^2$ and $k\in\ZZ$, define cones 
    \[\cC_k(\lambda)
      :=
      \begin{cases}
        \{(\lambda_1-r,\lambda_2+s):r,s\in\RR_{\ge0}\} & \text{if $k=2j$;}\\
        \{(\lambda_1+r,\lambda_2-s):r,s\in\RR_{\ge0}\} & \text{if $k=2j+1$.}
      \end{cases}
    \]
    The \emph{dominance region} $\cP(\lambda)$ is the intersection  $\bigcap_{k\in\ZZ}\phi_k^{-1}\cC_k(\phi_k\lambda)$.
    When $\mu\in\cP(\lambda)$, we say \emph{$\lambda$ dominates $\mu$}.
  \end{definition}

@@

  By an eigenvector of a piecewise-linear map $\phi$, we will mean a vector $\lambda$ such that for each $k>0$, there exists a scalar $\nu_k$ so that $\phi^k(\lambda)=\nu_k\lambda$.
  \begin{lemma}
    Any nonzero eigenvector of $\phi_2$ is a positive multiple of one of the vectors $\big(2b,-bc\pm\sqrt{bc(bc-4)}\big)$.
  \end{lemma}
  \begin{proof}
    First observe that the equation $\phi_2^k(\lambda)=\nu_k\lambda$ cannot be satisfied for all $k>0$ unless $\lambda_1\ge 0$ and $c\lambda_1+\lambda_2\ge 0$.
    In this region, $\phi_2$ is linear with eigenvalue $\nu$ satisfying $\nu^2-(bc-2)\nu+1=0$, i.e. $\nu=\frac{bc-2\pm\sqrt{bc(bc-4)}}{2}$.
    We thus require 
    \[\frac{bc-2\pm\sqrt{bc(bc-4)}}{2}\lambda_1=(bc-1)\lambda_1+b\lambda_2 \qquad\text{and}\qquad \frac{bc-2\pm\sqrt{bc(bc-4)}}{2}\lambda_2= -c\lambda_1-\lambda_2,\]
    or equivalently
    \[\frac{-bc\pm\sqrt{bc(bc-4)}}{2b}\lambda_1=\lambda_2 \qquad\text{and}\qquad \frac{-bc\mp\sqrt{bc(bc-4)}}{2c}\lambda_2=\lambda_1.\]
    As these represent the same relationship, the result immediately follows by inspection.
  \end{proof}

@@

  Here we explicitly compute the dominance regions $\cP(\lambda)$.
  When $\lambda$ is not imaginary, the dominance region is particularly simple.
  \begin{lemma}
    If $\lambda\in\ZZ^2\setminus\cI$, then $\cP(\lambda)=\{\lambda\}$.
  \end{lemma}
  \begin{proof}
    Any such $\lambda$ is the $\bfg$-vector of a cluster monomial, say $x_k^{a_1}x_{k+1}^{a_2}$.
    Then $\phi_{k-2}^{-1}\cC_{k-2}(\phi_{k-2}\lambda) \cap \phi_k^{-1}\cC_k(\phi_k\lambda)$ is precisely $\{\lambda\}$. 
    \sayDR{Check this, I am definitely convinced it works when $k$ is even.}
  \end{proof}

>>>>

  Fix integers $b,c>0$.  The cluster algebra $\cA(b,c)$ is the $\ZZ$-subalgebra of $\QQ(x_0,x_1)$ generated by \emph{cluster variables} $x_m$, $m\in\ZZ$, defined recursively by
  \begin{equation}
    x_{m-1}x_{m+1}=\begin{cases} x_m^b+1 & \text{if $m$ is even;}\\ x_m^c+1 & \text{if $m$ is odd.} \end{cases}
  \end{equation}
  By the Laurent Phenomenon \cite{fomin-zelevinsky}, each $x_m$ is actually an element of $\ZZ[x_k^{\pm1},x_{k+1}^{\pm1}]$ for any $k\in\ZZ$.
  Moreover, these Laurent polynomials are known to have positive coefficients \cite{LLZ,lee-schiffler,GHKK}.

  Our goal in this note is to explicitly compute the possible bases which are pointed with respect to the embedding $\cA(b,c)\into\ZZ[x_k^{\pm1},x_{k+1}^{\pm1}]$ for every $k\in\ZZ$.
  That is, a basis element with \emph{$\bfg$-vector} $\lambda\in\ZZ^2$ when expanded in terms of $\{x_0,x_1\}$ can be written in the form
  \begin{equation}
    \label{eq:pointed}
    x_0^{\lambda_1}x_1^{\lambda_2}\sum\limits_{a_1,a_2 \ge 0} c^\lambda_{a_1,a_2} x_0^{-ba_1} x_1^{ca_2}
  \end{equation}
  with $c^\lambda_{0,0}=1$.
  Being \emph{pointed} means an analogous structure reproduces after expanding in terms of any pair $\{x_k,x_{k+1}\}$.
  %The collection of all $\bfg$-vectors $\lambda\in\ZZ^2$ parametrize the elements of any pointed basis.

  Important examples of pointed elements are the \emph{cluster monomials} of $\cA(b,c)$, i.e. the elements that can be written in the form $x_k^{a_1}x_{k+1}^{a_2}$ for some $k\in\ZZ$ and $a_1,a_2$ non-negative integers.
  Indeed, by [someresult of Qin] cluster monomials are part of any pointed basis in any cluster algebra.
  \say{Check if this needs full rank}

@@

  A number of bases for rank two cluster algebras with various defining properties have been constructed.
  A natural starting basis is the so-called \emph{greedy basis} in which all $c^\lambda_{a_1,a_2}$ are non-negative and these coefficients are chosen to be minimal.
  It is known that this choice coincides with the theta basis \cite{GHKK, CGMMRSW}.
  The \emph{standard monomial bases} of $\cA(b,c)$ come from natural identifications with the subalgebras of $\ZZ[x_0^{\pm1},x_1^{\pm1}]$ generated by $x_{k-1}$, $x_k$, $x_{k+1}$, and $x_{k+2}$ for $k\in\ZZ$.
  However, these bases depend on $k$ and the triangular basis is defined by a triangular transition with the stipulation that the basis does not depend on the choice of starting cluster.
  It would be useful to explicitly (or combinatorially) compute the transitions between these bases.
  However, explicit calculations reveal that the $\bfg$-vectors actually appearing in these transitions are far smaller than the entire dominance polygon.

@@

  Given a Laurent polynomial in $\ZZ[x_0^{\pm1},x_1^{\pm1}]$, its \emph{support} is the set of its exponent vectors inside $\ZZ^2$.
  \begin{theorem}
    \label{th:maximum support}
    For a $\bfg$-vector $\lambda=(\lambda_1,\lambda_2)\in\ZZ^2$ the support for a pointed basis vector $x_\lambda$ is contained in the (possibly degenerate) lattice quadrilateral $\cS_\lambda$ given as follows:
    \begin{enumerate}
      \item If $\lambda$ lies outside of $\cI$, there are the following possibilities:
        \begin{enumerate}
          \item If $0\le\lambda_1,\lambda_2$, then $\cS_\lambda$ is just the point $\lambda$.
          \item If $\lambda_2 < 0$ and $0\le\lambda_1+b\lambda_2$, then $\cS_\lambda$ is the segment joining $\lambda$ and $(\lambda_1+b\lambda_2,\lambda_2)$.
          \item If $\lambda_1 < 0$ and $0\le\lambda_2$, then $\cS_\lambda$ is the segment joining $\lambda$ and $(\lambda_1,-c\lambda_1+\lambda_2)$.
          \item If $\lambda_1,\lambda_2 < 0$, then $\cS_\lambda$ has corner vertices $\lambda$, $(\lambda_1+b\lambda_2,\lambda_2)$, $\big(\lambda_1+b\lambda_2,-c\lambda_1-(bc-1)\lambda_2\big)$, and $(\lambda_1,-c\lambda_1+\lambda_2)$.
          \item If $\lambda_1 > 0$, $\lambda_1+b\lambda_2 < 0$, and $-c\lambda_1-(bc-1)\lambda_2\le 0$, then $\cS_\lambda$ has corner vertices $\lambda$, $(\lambda_1+b\lambda_2,\lambda_2)$, $\big(\lambda_1+b\lambda_2,-c\lambda_1-(bc-1)\lambda_2\big)$, and $\big((bc+1)\lambda_1-b^2c\lambda_2,-c\lambda_1-(bc-1)\lambda_2\big)$.
          \item If $\lambda_1 > 0$, $\lambda_1+b\lambda_2 < 0$, and $0 < -c\lambda_1-(bc-1)\lambda_2$, then $\cS_\lambda$ has corner vertices $\lambda$, $(\lambda_1+b\lambda_2,\lambda_2)$, $\big(\lambda_1+b\lambda_2,-c\lambda_1-(bc-1)\lambda_2\big)$, and $(0,0)$.
        \end{enumerate}
      \item If $\lambda$ lies inside of $\cI$, then $\cS_\lambda$ has corner vertices $\lambda$, $(\lambda_1+b\lambda_2,\lambda_2)$, $\big(\lambda_1+b\lambda_2,-c\lambda_1-(bc-1)\lambda_2\big)$, and $\frac{bc+\sqrt{bc(bc-4)}}{2\sqrt{bc(bc-4)}}\big(2\lambda_1+b\lambda_2,-c\lambda_1-(bc-2)\lambda_2\big)$.
    \end{enumerate}
  \end{theorem}

@@

\section{$\bfg$-vector Mutations}
  We begin by studying transformations of $\RR^2$ which determine the change of $\bfg$-vectors when expanding a pointed expression of the form~\eqref{eq:pointed} in terms of a cluster $\{x_k,x_{k+1}\}$.
  Write $\phi_0:\RR^2\to\RR^2$ for the identity map and define piecewise-linear maps $\phi_{\pm 1}:\RR^2\to\RR^2$ as follows:
  \begin{equation}
    \phi_1(\lambda)
    =
    \begin{cases} 
      (-\lambda_1,c\lambda_1+\lambda_2) & \text{if $\lambda_1 \ge 0$;}\\
      (-\lambda_1,\lambda_2) & \text{if $\lambda_1 < 0$;}
    \end{cases}
    \qquad
    \phi_{-1}(\lambda)
    =
    \begin{cases} 
      (\lambda_1,-\lambda_2) & \text{if $\lambda_2 \ge 0$;}\\
      (\lambda_1+b\lambda_2,-\lambda_2) & \text{if $\lambda_2 < 0$.}
    \end{cases}
  \end{equation}
  For $k\in\ZZ$ with $|k|>1$, define piecewise-linear maps
  \[\phi_k
    :=
    \begin{cases}
      (\phi_{-1}^{-1}\phi_1)^j & \text{if $k=2j$;}\\
      \phi_1(\phi_{-1}^{-1}\phi_1)^j & \text{if $k=2j+1$.}\\
    \end{cases}
  \]
  These determine the dominance region as follows.
  \begin{definition}
    For $\lambda\in\ZZ^2$ and $k\in\ZZ$, define cones 
    \[\cC_k(\lambda)
      :=
      \begin{cases}
        \{(\lambda_1-r,\lambda_2+s):r,s\in\RR_{\ge0}\} & \text{if $k=2j$;}\\
        \{(\lambda_1+r,\lambda_2-s):r,s\in\RR_{\ge0}\} & \text{if $k=2j+1$.}
      \end{cases}
    \]
    The \emph{dominance region} $\cP(\lambda)$ is the intersection  $\bigcap_{k\in\ZZ}\phi_k^{-1}\cC_k(\phi_k\lambda)$.
    When $\mu\in\cP(\lambda)$, we say \emph{$\lambda$ dominates $\mu$}.
  \end{definition}

@@

  By an eigenvector of a piecewise-linear map $\phi$, we will mean a vector $\lambda$ such that for each $n>0$, there exists a scalar $\nu_n$ so that $\phi^n(\lambda)=\nu_n\lambda$.
  \begin{lemma}
    Any nonzero eigenvector of $\phi_2$ is a positive multiple of one of the vectors $\big(2b,-bc\pm\sqrt{bc(bc-4)}\big)$.
  \end{lemma}
  \begin{proof}
    First observe that the equation $\phi_2^n(\lambda)=\nu_n\lambda$ cannot be satisfied for all $n>0$ unless $\lambda_1\ge 0$ and $c\lambda_1+\lambda_2\ge 0$.
    In this region, $\phi_2$ is linear with eigenvalue $\nu$ satisfying $\nu^2-(bc-2)\nu+1=0$, i.e. $\nu=\frac{bc-2\pm\sqrt{bc(bc-4)}}{2}$.
    We thus require 
    \[\frac{bc-2\pm\sqrt{bc(bc-4)}}{2}\lambda_1=(bc-1)\lambda_1+b\lambda_2 \qquad\text{and}\qquad \frac{bc-2\pm\sqrt{bc(bc-4)}}{2}\lambda_2= -c\lambda_1-\lambda_2,\]
    or equivalently
    \[\frac{-bc\pm\sqrt{bc(bc-4)}}{2b}\lambda_1=\lambda_2 \qquad\text{and}\qquad \frac{-bc\mp\sqrt{bc(bc-4)}}{2c}\lambda_2=\lambda_1.\]
    As these represent the same relationship, the result immediately follows by inspection.
  \end{proof}

@@

  Here we explicitly compute the dominance regions $\cP(\lambda)$.
  When $\lambda$ is not imaginary, the dominance region is particularly simple.
  \begin{lemma}
    If $\lambda\in\ZZ^2\setminus\cI$, then $\cP(\lambda)=\{\lambda\}$.
  \end{lemma}
  \begin{proof}
    Any such $\lambda$ is the $\bfg$-vector of a cluster monomial, say $x_k^{a_1}x_{k+1}^{a_2}$.
    Then $\phi_{k-1}^{-1}\cC_{k-1}(\phi_{k-1}\lambda) \cap \phi_{k+1}^{-1}\cC_{k+1}(\phi_{k+1}\lambda)$ is precisely $\{\lambda\}$. 
    \sayDR{Check this, I am definitely convinced it works when $k$ is odd.}
  \end{proof}
